Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2017-09-11 cabios-template/draft.tex
last changes

\newpage
\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge severely slowing down genomic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} and \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Genome-wide datasets produced for association studies have dramatically increased in size over the past years. A range of software and data formats have been developed to perform essential pre-processing steps and data analysis, often optimizing each of these steps within a dedicated implementation. This diverse and extremely rich software environment has been of tremendous benefit for the genetic community. However, it has two limitations: analysis pipelines become very complex and researchers have limited access to diverse analysis tools due to growing data sizes.

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}. Similarly, quality control software for genome-wide analysis have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}. Computation of principal components (PCs) of genotypes is commonly performed to account for population stratification in association studies and there are several software available, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}. Then, implementation of GWAS analyses also depends on the genotype format, e.g. ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple genetic variants \cite[]{Vilhjalmsson2015,Euesden2015}. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this might be a brake on data exploration.
%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis that have been already implemented. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.

@@

We developed two R packages, bigstatsr and bigsnpr, that integrate the most efficient algorithms for the pre-processing and analysis of large-scale genomic data while using memory-mapping. Package bigstatsr implements many statistical tools for several types of ``big.matrix'' objects (raw, char, short, integer, float and double). This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of ``big.matrix'' object to store the genotypes. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (pruning, testing, plotting).

We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores. We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.

@@

The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory. They use what we could call an hybrid solution: memory-mapping. Memory-mapping is used to access data, possibly stored on disk, as if it were in memory. This solution is made available within R through an object called ``big.matrix'', available in R package bigmemory \cite[]{Kane2013}.

We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{RSpectra2016,Zeng2017}.

Usually, memory-mapping provides seamless and faster access than standard read/write operations. When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. All this is managed by the Operating System so that it is seamless and efficient. It means that if you use the same chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on. Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.

@@

Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).

Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}). The first function is a wrapper to PLINK and Beagle which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R \cite[]{Browning2009}. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This algorithm doesn't use phasing and is very fast, as previous Machine Learning Approaches for genetic imputation \cite[]{Wang2012}. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}. XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices. For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets. The training set is used to build a model for predicting missing data. The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of individuals that have been wrongly imputed for that particular SNP.

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).

The function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. FlashPCA2 and bigstatsr use the same PCA algorithm called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the ``big.matrix'' format, which enables parallel computations.

Fast algorithms for thinning SNPs  similar to algorithms provided in PLINK have been developed. For instance, thinning is mandatory when computing PCs of a genotype matrix \cite[]{Abdellaoui2013}. There are at least 3 different options to thin SNPs based on Linkage Disequilibrium. The first option is known as pruning, which is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning.

A variant of pruning is clumping. Clumping is useful if a statistic is available to sort the SNPs by importance, e.g. association with a phenotype, and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest. Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}. However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
Statistical tests based on linear and logistic regressions are also available. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.

The R packages also implement functions to compute Polygenic Risk Scores. 
First, they implement the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations. Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). Finally, a polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Dudbridge2013,Chatterjee2013,Golan2014}.

Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are linear and logistic regressions and Support Vector Machine (SVM). These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Tibshirani1996,Zou2005,Friedman2010}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).

@@

In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.

For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset. \label{sec:rep}

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT). Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. 
The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with the XGBoost method. If we used a standard R matrix to store the genotypes, this data would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB only.

We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer. The GWAS analysis was composed of three main steps. First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs. Finally, on the whole dataset, we made a linear association test for each SNP, using the 10 first PCs as covariables.

@@

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-10}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}

>>>>

\newpage
\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge severely slowing down genomic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} \& \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

Genome-wide datasets produced for association studies have dramatically increased in size over the past years. A range of software and data formats have been developed to perform essential pre-processing steps and data analysis, often optimizing each of these steps within a dedicated implementation. This diverse and extremely rich software environment has been of tremendous benefit for the genetic community. However, it has two limitations: analysis pipelines are becoming very complex and researchers have limited access to diverse analysis tools due to growing data sizes.

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}. Similarly, quality control software for genome-wide analysis have been developed such as PLINK and the Bioconductor package GWASTools \cite[]{Gogarten2012}. 
There are also several software for the computation of principal components (PCs) of genotypes, commonly performed to account for population stratification in association studies, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Abraham2014a,Abraham2016a,Galinsky2016,Price2006}. 
Then, implementation of GWAS analyses also depends on the data format and model analyzed. For example, the software ProbABEL \cite[]{Aulchenko2010} or SNPTEST \cite[]{Marchini2010} can handle dosage data, while PLINK version 1 is limited to best guess genotypes because of its input file format. 
Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred \cite[]{Vilhjalmsson2015} and PRSice \cite[]{Euesden2015}, which provide prediction for quantitative traits or disease risks based on multiple genetic variants. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this might be a brake on data exploration.
%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis that have been already implemented. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping has been available for several years in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}. 
We provide a similar format as filebacked ``big.matrix'' objects that we called ``Filebacked Big Matrices (FBMs)''.
Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.

@@

We developed two R packages, bigstatsr and bigsnpr, that integrate the most efficient algorithms for the pre-processing and analysis of large-scale genomic data while using memory-mapping. Package bigstatsr implements many statistical tools for several types of FBMs (unsigned char, unsigned short, integer and double). This includes implementation of multivariate sparse linear models, Principal Component Analysis, matrix operations, and numerical summaries. The statistical tools developed in bigstatsr can be used for other types of data as long as they can be represented as matrices. Package bigsnpr depends on bigstatsr, using a special type of FBM object to store the genotypes, called ``FBM.code256''. Package bigsnpr implements algorithms which are specific to the analysis of SNP arrays, such as calls to external software for processing steps, I/O (Input/Output) operations from binary PLINK files, and data analysis operations on SNP data (thinning, testing, plotting). 
We use both a real case-control genomic dataset for Celiac disease and large-scale simulated data to illustrate application of the two R packages, including association study and computation of Polygenic Risk Scores. We also compare results from the two R packages with those obtained when using PLINK and EIGENSOFT, and report execution times along with the code to perform major computational tasks.

@@

The two R packages don't use standard read operations on a file nor load the genotype matrix entirely in memory. They use an hybrid solution: memory-mapping. Memory-mapping is used to access data, possibly stored on disk, as if it were in memory. This solution is made available within R through the BH package, providing access to Boost C++ Header Files\footnote{http://www.boost.org/}.

We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008} and of the R package BEDMatrix\footnote{https://github.com/QuantGen/BEDMatrix} which provides memory-mapping directly for binary PLINK files. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso and RSpectra \cite[]{RSpectra2016,Zeng2017}.

Memory-mapping provides transparent and faster access than standard read/write operations. 
When an element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. 
When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. 
All this is managed by the Operating System so that it is seamless and efficient. 
It means that if the same chunks of data are used repeatedly, it will be very fast the second time they are accessed, the third time and so on. 
Of course, if the memory size of the computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.

@@

Package bigsnpr currently takes as input a variety of formats (e.g. vcf, bed/bim/fam, ped/map). However, it uses PLINK for conversion to bed/bim/fam format and for Quality Control (QC) of the data, so that we first provide R functions that use system calls to PLINK for the conversion and QC steps (Figure~\ref{fig:qc}).
Then, fast read/write operations from/to bed/bim/fam PLINK files are implemented. PLINK files are then read into a ``bigSNP'' object, which contains the genotype Filebacked Big Matrix (FBM), a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``FBM''.

We developed a special FBM object, called ``FBM.code256'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).

We also provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}). The first function is a wrapper to PLINK and Beagle \cite[]{Browning2009} which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This algorithm is based on Machine Learning approaches for genetic imputation \cite[]{Wang2012} and doesn't use phasing, thus allowing for a dramatic decrease in computation time. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}. 
XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase, making it well suited for imputing genotype matrices. Systematically, for each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets. The training set is used to build a model for predicting missing data. The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of genotypes that have been wrongly imputed for that particular SNP.

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}). While the function based on eigenvalue decomposition is at least quadratic in the smallest dimension, the function based on randomized projections runs in linear time in all dimensions \cite[]{Lehoucq1996}. Package bigstatsr use the same PCA algorithm as FlashPCA2 called Implicitly Restarted Arnoldi Method (IRAM), which is implemented in R package RSpectra. The main difference between the two implementations is that FlashPCA2 computes vector-matrix multiplications with the genotype matrix based on the binary PLINK file whereas bigstatsr computes these multiplications based on the FBM format, which enables parallel computations and easier subsetting.

SNP thinning improves ascertainment of population structure with PCA \cite[]{Abdellaoui2013}. There are at least 3 different approaches to thin SNPs based on Linkage Disequilibrium, two of them named pruning and clumping, address SNPs in LD close to each others because of recombination events, while the third one address long-range regions with a complex LD pattern due to other biological events such as inversions \cite[]{Price2008}. 
First, pruning, the most naive approach, is an algorithm that sequentially scan the genome for nearby SNPs in LD, performing pairwise thinning based on a given threshold of correlation.
A variant of pruning is clumping. Clumping is useful if a statistic is available to sort the SNPs by importance, e.g. association with a phenotype, and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest. Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.

As mentioned above, the third approach that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}. 
However, the pattern of LD might be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
We also implement statistical tests based on linear and logistic regressions. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $$\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $K$ is the number of principal components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $$\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L,$$ and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.

The R packages also implement functions to compute Polygenic Risk Scores using two approaches. 
The first method is the widely-used Pruning + Thresholding (P+T) model based on univariate GWAS summary statistics as described in previous equations. Under the P+T model, a coefficient of regression is learned independently for each SNP along with a corresponding p-value. The SNPs are first clumped (P) so that there remains only SNPs that are weakly correlated with each other. Thresholding (T) consists in removing SNPs that are under a certain level of significance (P-value threshold to be determined). A polygenic risk score is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding regression coefficients \cite[]{Chatterjee2013,Dudbridge2013,Golan2014}. 
The second approach doesn't use univariate summary statistics but instead train a multivariate model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are linear and logistic regressions and Support Vector Machine (SVM). These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Friedman2010,Tibshirani1996,Zou2005}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models and to use these algorithms on the special type of FBM called ``FBM.code256'' used in bigsnpr.

@@

In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.
For computation times comparison, we replicated individuals in the Celiac dataset 5 and 10 times in order to increase sample size while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also through replication of the Celiac dataset. \label{sec:rep}

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT). Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. We compare performances on two computers, a desktop computer with 64GB of RAM and 12 cores (6 physical cores), and a laptop with only 8GB of RAM and 4 cores (2 physical cores). For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. 
The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing the first individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\%. The 0.04\% missing genotype values were imputed with the XGBoost method. If we used a standard R matrix to store the genotypes, this data would require 32GB of memory. On the disk, the ``.bed'' file requires 1GB and the ``.bk'' file (storing the FBM) requires 4GB.

We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of the Celiac genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while adjusting for PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer. The GWAS analysis was composed of three main steps. First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs. Finally, we performed a linear association test in the complete 500K dataset for each of the 1M SNPs, using the 10 first PCs as covariables.

@@

\newpage
\begin{figure}[!h]
\centerline{\includegraphics[width=\textwidth]{celiac-gwas-cut}}
\caption{Manhattan plot of the celiac disease cohort produced by package bigsnpr. Some SNPs in chromosome 6 have p-values smaller than the $10^{-30}$ threshold used for vizualisation purposes.}\label{fig:gwas}
\end{figure}
